### 2020-10-11 conclusion.tex
Messing around with schedulers.

\paragraph{On Curry-Howard correspondences}
The Curry-Howard correspondence between logic and types laid the foundation for functional programming~\cite{wadler12}. With the rise of linear logic~\cite{girard87} a Curry-Howard correspondence was established between linear logic and the linear $\pi$-calculus~\cite{abramsky94,bellinscott94}. With the rise of session types, a new correspondence between linear logic and the $\pi$-calculus was proposed first by Caires and Pfenning~\cite{cairespfenning10}, for dual intuitionistic linear logic (DILL), and later by Wadler~\cite{wadler12} for classical linear logic (CLL). The underlying language in both correspondences satisfy deadlock freedom by design, by forbidding cyclic connections via typing. Dardha and Gay follow Padovani's work~\cite{padovani14} and use priorities in CP to define a new calculus which is deadlock-free by design, based on classical linear logic~\cite{girard87}  and the Curry-Howard correspondence with session types~\cite{wadler12}. Dardha and P\'{e}rez~\cite{dardhaperez15} compare Kobayashi-style typing and CLL typing, and prove that CLL corresponds to Kobayashi's system with the restriction that only single cuts are allowed. Balzer \emph{et al.} study sharing~\cite{balzerpfenning17} and later on deadlock freedom~\cite{balzertoninho19} in session-typed $\pi$-calculus based on DILL. In~\cite{balzerpfenning17} types are either linear or shared with modal operators connecting them, which allows sharing, but not deadlock freedom. Later on in~\cite{balzertoninho19} types are enhanced with information regarding the order of acquiring resources, which guarantees deadlock freedom.

>>>>

\paragraph{On Curry-Howard correspondences}
The Curry-Howard correspondence between logic and types laid the foundation for functional programming~\cite{wadler15}. With the rise of linear logic~\cite{girard87} a Curry-Howard correspondence was established between linear logic and the linear $\pi$-calculus~\cite{abramsky94,bellinscott94}. With the rise of session types, a new correspondence between linear logic and the $\pi$-calculus was proposed first by Caires and Pfenning~\cite{cairespfenning10}, for dual intuitionistic linear logic (DILL), and later by Wadler~\cite{wadler15} for classical linear logic (CLL). The underlying language in both correspondences satisfy deadlock freedom by design, by forbidding cyclic connections via typing. Dardha and Gay follow Padovani's work~\cite{padovani14} and use priorities in CP to define a new calculus which is deadlock-free by design, based on classical linear logic~\cite{girard87}  and the Curry-Howard correspondence with session types~\cite{wadler15}. Dardha and P\'{e}rez~\cite{dardhaperez15} compare Kobayashi-style typing and CLL typing, and prove that CLL corresponds to Kobayashi's system with the restriction that only single cuts are allowed. Balzer \emph{et al.} study sharing~\cite{balzerpfenning17} and later on deadlock freedom~\cite{balzertoninho19} in session-typed $\pi$-calculus based on DILL. In~\cite{balzerpfenning17} types are either linear or shared with modal operators connecting them, which allows sharing, but not deadlock freedom. Later on in~\cite{balzertoninho19} types are enhanced with information regarding the order of acquiring resources, which guarantees deadlock freedom.

### 2020-10-11 introduction.tex
Messing around with schedulers.

Things to discuss in the intro:

\begin{itemize}
\item GV is a session-typed functional language.
\item
  The benefits of using a functional language as opposed to a process calculus are that:
  \begin{itemize}
  \item 
    it supports higher-order functions (and therefore abstraction) not usually present in process calculi; and
  \item
    it allows us to derive extensions of the communication fragment of the language via well-understood extensions of the functional fragment, i.e., deriving internal/external choice from sum types.
  \end{itemize}
\item
  The benefits of using GV over other session-typed functional languages are that GV has strong ties to linear logic, via its relation to CP~\cite{wadler12}, and consequently it has strong formal properties, e.g., deadlock freedom.
\item
  Historically, calculi in the GV family~\cite{wadler12,lindleymorris15} have achieved deadlock freedom via a syntactic restriction, i.e., by combining channel creation and thread creation into a single operation, called fork, corresponding to the logical cut in CP, itself the combination of a name restriction and a parallel composition.
  Unfortunately, this is overly restrictive, as it limits the possible communication structures to trees. The calculus GV is based on, on the other hand, did not satisfy deadlock freedom~\cite{gayvasconcelos10}.
\item
  Recent developments in CP have led to various approaches to decoupling these constructs, either while maintaining the strong correspondence to logic, as in Hypersequent CP~\cite[HCP]{kokkemontesi19popl,kokkemontesi19tlla}, or by weakening the correspondence to logic in exchange for a more expressive language, as in Priority CP~\cite[PCP]{dardhagay18}.
\item
  PCP decouples CP's cut into separate constructs for name restriction and parallel composition, and restores deadlock freedom by adding priorities \`{a} la~\cite{kobayashi06}.
\item
  Following the example set by PCP, we present Priority GV (PGV), a~variant of GV which decouples channel creation from thread creation, and restores deadlock freedom by adding priorities.
\item
  We demonstrate the correspondence between PGV and PCP by proving operational correspondence for a translation from PCP to PGV.
\end{itemize}

>>>>

We introduce Priority GV (PGV), a~session-typed concurrent $\lambda$-calculus.

\begin{itemize}
\item
  Historically, calculi in the GV family~\cite{wadler15,lindleymorris15} have achieved deadlock freedom via a syntactic restriction, i.e., by combining channel creation and thread spawning into a single operation, called fork, corresponding to the logical cut in CP, itself the combination of a name restriction and a parallel composition.
  Unfortunately, this is overly restrictive, as it limits the possible communication structures to trees. The calculus GV is based on, on the other hand, did not satisfy deadlock freedom~\cite{gayvasconcelos12}.
\item
  Recent developments in CP have led to various approaches to decoupling these constructs, either while maintaining the strong correspondence to logic, as in Hypersequent CP~\cite[HCP]{kokkemontesi19popl,kokkemontesi19tlla}, or by weakening the correspondence to logic in exchange for a more expressive language, as in Priority CP~\cite[PCP]{dardhagay18}.
\item
  PCP decouples CP's cut into separate constructs for name restriction and parallel composition, and restores deadlock freedom by adding priorities \`{a} la~\cite{kobayashi06}.
\item
  Following the example set by PCP, we present Priority GV (PGV), a~variant of GV which decouples channel creation from thread creation, and restores deadlock freedom by adding priorities.
\item
  We demonstrate the correspondence between PGV and PCP by proving operational correspondence for a translation from PCP to PGV.
\item
  The benefits of using a functional language as opposed to a process calculus are that:
  \begin{itemize}
  \item 
    it supports higher-order functions (and therefore abstraction) not usually present in process calculi; and
  \item
    it allows us to derive extensions of the communication fragment of the language via well-understood extensions of the functional fragment, i.e., deriving internal/external choice from sum types.
  \end{itemize}
\item
  The benefits of using GV over other session-typed functional languages are that GV has strong ties to linear logic, via its relation to CP~\cite{wadler15}, and consequently it has strong formal properties, e.g., deadlock freedom.
\end{itemize}

### 2020-10-11 priority-cp.tex
Messing around with schedulers.

\paragraph*{Terms}
We let $\tm{P}, \tm{Q}$ range over processes, produced by the following grammar.
\[
\begin{array}[t]{lcl}
  \tm{P}, \tm{Q}
  & \coloneqq & \tm{\link{x}{y}}
         \sep   \tm{\res{x}{y}{P}}
         \sep   \tm{(\ppar{P}{Q})}
         \sep   \tm{\halt}
  \\   & \sep & \tm{\send{x}{y}{P}}
         \sep   \tm{\close{x}{P}}
         \sep   \tm{\recv{x}{y}{P}}
         \sep   \tm{\wait{x}{P}}
  \\   & \sep & \tm{\inl{x}{P}}
         \sep   \tm{\inr{x}{P}}
         \sep   \tm{\offer{x}{P}{Q}}
         \sep   \tm{\absurd{x}}
\end{array}
\]
Process $\tm{\link{x}{y}}$ links channels $x$ and $y$ and forwards communication from one to the other. $\tm{\res{x}{y}{P}}$, $\tm{(\ppar{P}{Q})}$ and $\tm{\halt}$ denote respectively the restriction processes where channel endpoints $\tm x$ and $\tm y$ are bound together and with scope $\tm P$, the parallel composition of processes $\tm{P}$ and $\tm{Q}$ and the terminated process.

Processes $\tm{\send{x}{y}{P}}$ and $\tm{\recv{x}{y}{P}}$ send or receive over channel $\tm x$ a value $\tm y$ and proceed as process $\tm P$. $\tm{\close{x}{P}}$ and $\tm{\wait{x}{P}}$ send and receive an empty value -- denoting the closure of channel $\tm x$, and continue as $\tm P$.

Processes $\tm{\inl{x}{P}}$ and $\tm{\inr{x}{P}}$ make a left and right choice, respectively and proceed as process $\tm P$. $\tm{\offer{x}{P}{Q}}$ dually offers both left and right branches, with continuations $\tm P$ and $\tm Q$. $\tm{\absurd{x}}$ denotes the empty offer.

Finally, we define \emph{unbound} send as syntactic sugar:
\[
  \tm{\usend{x}{y}{P}}\elabarrow\tm{\send{x}{z}{(\ppar{\link{y}{z}}{P})}}
\]

@@

Typing rules for our version of PCP are given in \Cref{fig:pcp-typing}. A typing judgement $\seq{P}{\Gamma}$ states that "process $\tm P$ is well typed under the typing context $\ty \Gamma$".

Typing rules make use of priorities, defined in the previous section. Priorities are based on \emph{obligations/capabilities} used by Kobayashi~\cite{kobayashi06}, and simplified to single priorities following Padovani~\cite{padovani14}. Recalling PCP \cite{dardhagay18}, priorities obey the following two laws:
\begin{enumerate}
  \item [(i)] an action of priority $\cs o$ must be prefixed only by actions of priorities \emph{strictly smaller} than $\cs o$.
  \item [(ii)] communication requires \emph{equal} priorities for the dual actions.
\end{enumerate}
\textsc{T-Link} states that the link process $\tm{\link{x}{y}}$ is well typed under channels $\tm x$ and $\tm y$ having dual types, respectively $\ty A$ and $\ty {\co A}$. \textsc{T-Res} sates that the restriction process $\tm{\res{x}{y}{P}}$ is well typed under typing context $\ty\Gamma$ if process $\tm P$ is well typed in $\ty\Gamma$ augmented with channel endpoints $\tm x$ and $\tm y$ having dual types, respectively $\ty A$ and $\ty {\co A}$. \textsc{T-Par} states that the parallel composition of processes $\tm P$ and $\tm Q$ is well typed under the disjoint union of their respective typing contexts. \textsc{T-Halt} states that the terminated process $\tm{\halt}$ is well typed in the empty context.

\textsc{T-Send} and \textsc{T-Recv} state that the sending and receiving of a bound name $\tm y$ over a channel $\tm x$ is well typed under $\ty\Gamma$ and $\tm x$ of type $\tytens[\cs{o}]{A}{B}$, respectively ${\typarr[\cs{o}]{A}{B}}$. Priority $\cs o$ is the smallest among all priorities of the types used by the output or input process, captured by the side condition $\cs{o}<\minpr(\ty{\Gamma},\ty{A},\ty{B})$.

Rules \textsc{T-Close} and \textsc{T-Wait} type the closure of channel $\tm x$ and are in the same lines as the previous two rules, requiring that the priority of channel $\tm x$ is the smallest among all priorities in $\tm\Gamma$.

\textsc{T-Select-Inl} and \textsc{T-Select-Inr} type respectively the left $\tm{\inl{x}{P}}$ and right $\tm{\inr{x}{P}}$ choice performed on channel $\tm x$. \textsc{T-Offer} and \textsc{T-Offer-Absurd} type the offering of a choice, or empty choice, on channel $\tm x$. In all the above rules the priority $\cs o$ of channel $\tm x$ is the smallest with respect to the typing context $\cs{o}<\minpr(\ty{\Gamma})$ and types involved in the choice $\cs{o}<\minpr(\ty{\Gamma},\ty{A},\ty{B})$.

\begin{example}[Milner's Cyclic Scheduler \Cref{fig:scheduler}]
  % 
  We will illustrate PCP with the cyclic scheduler~\cite{milner89}. The full typing for $\tm {Sched}$ is given in \cite{dardhagay18}.

  We will consider a simplified version: a set of 3 agents $\tm {A_0}, \tm {A_1}, \tm {A_2}$ is scheduled to perform a certain task in cyclic order, starting with agent $\tm {A_0}$. Agents $\tm {A_1}$ and $\tm {A_2}$ will send the result of some computation a collector process, respectively $\tm {P_1}$ and $\tm {P_2}$, before transmitting data further to the next agent in the cycle, being $\tm {A_{(i+1)\texttt{ mod } 3}}$. At the end of the round, agent $\tm {A_0}$ will send the final result of the computation to its collector $\tm {P_0}$.

  In line with the logical consistency requirements of the Curry-Howard correspondence, we define a finite version of Milner's scheduler, which executes one round of communication.
  $$
    \begin{array}{rcll}
      \tm {Sched} & \defeq & ..\tm {(\nu {a_i}{b_i})}..\tm {(\nu {c_i}{d_{{(i+1)\texttt{ mod } n}}})} \tm {\big(A_0 \parallel A_1 \parallel A_2 \parallel P_0 \parallel P_1 \parallel P_2 \big)}\\
      \tm{A_0} & \defeq & \tm{\send{c_0}{n_0}{\recv{d_0}{x_0}{\send{a_0}{m_0}{close_0}}}}\\
      \tm{A_i} & \defeq & \tm{\recv{d_i}{x_i}{\send{a_i}{m_i}{\send{c_i}{n_i}{close_i}}}} &~~ i \in \{1,2\}\\
      \tm {P_i} & \defeq & \tm{\recv{b_i}{y_i}{Q_i}} &~~ i \in \{0,1,2\}
    \end{array}
  $$

  In the above process, channels $n_i$ and $m_i$ for all $i$, denote data being transmitted, and for simplicity we let them be channels of a terminated type. Process $\tm{close_i}$ for all $i$ closes all channels in $\tm{A_i}$. In particular, they are a sequence of $\tm{\close{x}{P}}$ and $\tm{\wait{x}{P}}$ for all channels in the agent and terminating in $\tm{\halt}$.
\end{example}

\subsection{Operational Semantics}
\input{fig/pcp-operational-semantics}

@@

The reduction relation is given by a set of axioms and inference rules for context closure. Reduction occurs under restriction following \cite{dardhagay18} where channel endpoints are identified.
$\LabTirName{E-Link}$ reduces a parallel composition with a link into a substitution. $\LabTirName{E-Send}$ is the main communication rule, where send and receive processes sychronise and reduce to the corresponding continuations. $\LabTirName{E-Close}$ follows the previous rule and it closes the channel identified by endpoints $\tm x$ and $\tm y$. $\LabTirName{E-Select-Inl}$ and $\LabTirName{E-Select-Inr}$ are generalised versions of $\LabTirName{E-Send}$. They state respectively that a left and right selection synchronises with a choice offering and reduces to the corresponding continuations. The last three rules state that reduction is closed under restriction, parallel composition and structural congruence, respectively.

>>>>

\paragraph*{Terms}
We let $\tm{P}, \tm{Q}$ range over processes, produced by the following grammar.
\[
\begin{array}[t]{lcl}
  \tm{P}, \tm{Q}
  & \coloneqq & \tm{\link{x}{y}}
         \sep   \tm{\res{x}{y}{P}}
         \sep   \tm{(\ppar{P}{Q})}
         \sep   \tm{\halt}
  \\   & \sep & \tm{\send{x}{y}{P}}
         \sep   \tm{\close{x}{P}}
         \sep   \tm{\recv{x}{y}{P}}
         \sep   \tm{\wait{x}{P}}
  \\   & \sep & \tm{\inl{x}{P}}
         \sep   \tm{\inr{x}{P}}
         \sep   \tm{\offer{x}{P}{Q}}
         \sep   \tm{\absurd{x}}
\end{array}
\]
Process $\tm{\link{x}{y}}$ links channels $x$ and $y$ and forwards communication from one to the other. $\tm{\res{x}{y}{P}}$, $\tm{(\ppar{P}{Q})}$ and $\tm{\halt}$ denote respectively the restriction processes where channel endpoints $\tm{x}$ and $\tm{y}$ are bound together and with scope $\tm{P}$, the parallel composition of processes $\tm{P}$ and $\tm{Q}$ and the terminated process.

Processes $\tm{\send{x}{y}{P}}$ and $\tm{\recv{x}{y}{P}}$ send or receive over channel $\tm{x}$ a value $\tm{y}$ and proceed as process $\tm{P}$. $\tm{\close{x}{P}}$ and $\tm{\wait{x}{P}}$ send and receive an empty value -- denoting the closure of channel $\tm{x}$, and continue as $\tm{P}$.

Processes $\tm{\inl{x}{P}}$ and $\tm{\inr{x}{P}}$ make a left and right choice, respectively and proceed as process $\tm{P}$. $\tm{\offer{x}{P}{Q}}$ dually offers both left and right branches, with continuations $\tm{P}$ and $\tm{Q}$. $\tm{\absurd{x}}$ denotes the empty offer.

We write \emph{unbound} send as $\tm{\usend{x}{y}{P}}$, which is syntactic sugar for $\tm{\send{x}{z}{(\ppar{\link{y}{z}}{P})}}$. Alternatively, we could take $\tm{\usend{x}{y}{P}}$ as primitive, and let $\tm{\send{x}{y}{P}}$ be syntactic sugar for $\tm{\res{y}{z}{(\usend{x}{z}{P})}}$. CP takes \emph{bound} sending as primitive, as it is impossible to eliminate the top-level cut in terms such as $\tm{\res{y}{z}{(\usend{x}{z}{P})}}$, even with communting conversions. However, in our setting without commuting conversions and with more permissive normal forms, this is no longer an issue.

@@

Typing rules for our version of PCP are given in \Cref{fig:pcp-typing}. A typing judgement $\seq{P}{\Gamma}$ states that "process $\tm{P}$ is well typed under the typing context $\ty \Gamma$".

Typing rules make use of priorities, defined in the previous section. Priorities are based on \emph{obligations/capabilities} used by Kobayashi~\cite{kobayashi06}, and simplified to single priorities following Padovani~\cite{padovani14}. Recalling PCP \cite{dardhagay18}, priorities obey the following two laws:
\begin{enumerate}
  \item [(i)] an action of priority $\cs o$ must be prefixed only by actions of priorities \emph{strictly smaller} than $\cs o$.
  \item [(ii)] communication requires \emph{equal} priorities for the dual actions.
\end{enumerate}
\textsc{T-Link} states that the link process $\tm{\link{x}{y}}$ is well typed under channels $\tm{x}$ and $\tm{y}$ having dual types, respectively $\ty A$ and $\ty {\co A}$. \textsc{T-Res} sates that the restriction process $\tm{\res{x}{y}{P}}$ is well typed under typing context $\ty\Gamma$ if process $\tm{P}$ is well typed in $\ty\Gamma$ augmented with channel endpoints $\tm{x}$ and $\tm{y}$ having dual types, respectively $\ty A$ and $\ty {\co A}$. \textsc{T-Par} states that the parallel composition of processes $\tm{P}$ and $\tm{Q}$ is well typed under the disjoint union of their respective typing contexts. \textsc{T-Halt} states that the terminated process $\tm{\halt}$ is well typed in the empty context.

\textsc{T-Send} and \textsc{T-Recv} state that the sending and receiving of a bound name $\tm{y}$ over a channel $\tm{x}$ is well typed under $\ty\Gamma$ and $\tm{x}$ of type $\tytens[\cs{o}]{A}{B}$, respectively ${\typarr[\cs{o}]{A}{B}}$. Priority $\cs o$ is the smallest among all priorities of the types used by the output or input process, captured by the side condition $\cs{o}<\minpr(\ty{\Gamma},\ty{A},\ty{B})$.

Rules \textsc{T-Close} and \textsc{T-Wait} type the closure of channel $\tm{x}$ and are in the same lines as the previous two rules, requiring that the priority of channel $\tm{x}$ is the smallest among all priorities in $\tm\Gamma$.

\textsc{T-Select-Inl} and \textsc{T-Select-Inr} type respectively the left $\tm{\inl{x}{P}}$ and right $\tm{\inr{x}{P}}$ choice performed on channel $\tm{x}$. \textsc{T-Offer} and \textsc{T-Offer-Absurd} type the offering of a choice, or empty choice, on channel $\tm{x}$. In all the above rules the priority $\cs o$ of channel $\tm{x}$ is the smallest with respect to the typing context $\cs{o}<\minpr(\ty{\Gamma})$ and types involved in the choice $\cs{o}<\minpr(\ty{\Gamma},\ty{A},\ty{B})$.

\subsection{Example: Milner's Cyclic Scheduler}
We illustrate PCP with the cyclic scheduler~\cite{milner89}. The full typing for $\tm{\sched}$ is given in \cite{dardhagay18}.

We consider a simplified version: a set of 3 agents $\tm{\agent0}, \tm{\agent1}, \tm{\agent2}$ is scheduled to perform a certain task in cyclic order, starting with $\tm{\agent0}$. $\tm{\agent1}$ and $\tm{\agent2}$ send the result of some computation a collector process, respectively $\tm{\collector1}$ and $\tm{\collector2}$, before transmitting data further to the next agent in the cycle, being $\tm{\agent{(i+1)\bmod 3}}$. At the end of the round, $\tm{\agent0}$ send the final result of the computation to its collector $\tm{\collector0}$.

In line with the logical consistency requirements of the Curry-Howard correspondence, we define a finite version of Milner's scheduler, which executes one round of communication.
$$
\begin{array}{rcll}
  \tm{\sched}
  & \defeq
  & \tm{
    \res{x_0}{x'_0}{\res{x_1}{x'_1}{\res{x_2}{x'_2}{
    \res{\overrightarrow{y_0}}{\overleftarrow{y_1}}{
    \res{\overrightarrow{y_1}}{\overleftarrow{y_2}}{
    \res{\overrightarrow{y_2}}{\overleftarrow{y_0}}{}}}}}}
    }\\
  &
  & \tm{(\ppar{\agent0}{\ppar{\agent1}{\ppar{\agent2}{\ppar{\collector0}{\ppar{\collector1}{\collector2}}}}})}\\
  \tm{\agent0}
  & \defeq
  & \tm{\usend{\overrightarrow{y_0}}{d_0}{\recv{\overleftarrow{y_0}}{d'_0}{\usend{x_0}{d'_0}{
    \wait{\overleftarrow{y_0}}{\close{\overrightarrow{y_0}}{\close{x_0}{\halt}}}}}}} \\
  \tm{\agent{i}}
  & \defeq
  & \tm{\recv{\overleftarrow{y_i}}{d_i}{\usend{x_i}{d_i}{\recv{x_i}{d'_i}{\usend{\overrightarrow{y_i}}{d'_i}{
    \wait{\overleftarrow{y_i}}{\close{\overrightarrow{y_i}}{\close{x_i}{\halt}}}}}}}}
  & i \in \{1,2\} \\
  \tm{\collector{i}}
  & \defeq & \tm{\recv{x'_i}{d_i}{Q_i}}
  & i \in \{0,1,2\}
\end{array}
$$
In the above process, channels $\tm{d_i}$ and $\tm{d'_i}$ represent the data being transmitted. For simplicity, let them be channels of a terminated type. The processes $\tm{Q_i}$ represent the inner workings of the collector processes, each of which must eventually send back a response over $\tm{x'_i}$ and close the channel.
\subsection{Operational Semantics}
\input{fig/pcp-operational-semantics}

@@

The reduction relation is given by a set of axioms and inference rules for context closure. Reduction occurs under restriction following \cite{dardhagay18} where channel endpoints are identified.
$\LabTirName{E-Link}$ reduces a parallel composition with a link into a substitution. $\LabTirName{E-Send}$ is the main communication rule, where send and receive processes sychronise and reduce to the corresponding continuations. $\LabTirName{E-Close}$ follows the previous rule and it closes the channel identified by endpoints $\tm{x}$ and $\tm{y}$. $\LabTirName{E-Select-Inl}$ and $\LabTirName{E-Select-Inr}$ are generalised versions of $\LabTirName{E-Send}$. They state respectively that a left and right selection synchronises with a choice offering and reduces to the corresponding continuations. The last three rules state that reduction is closed under restriction, parallel composition and structural congruence, respectively.
